Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-03-06 manuscript/manuscript.qmd
New stuff in analysis

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the Covid-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. However, the abuse of emergency derogations by many governments raises concerns about the long-lasting impact of these sweeping measures on human rights. Does a robust civil society mitigate the adverse effects arising from international treaty derogations? Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run. Using data from the Varieties of Democracy PanDem dataset, the Oxford COVID-19 Government Response Tracker and the CoronaNet COVID-19 Government Response Event, we first assess the variation in human rights responses among governments during the pandemic. We then assess the role of civil society in keeping governments accountable and mitigating the effects of policies that violate human rights. Our results hold important implications for understanding government compliance with international treaties and shedding light on how states of emergency may directly and indirectly lead to democratic backsliding.
date: February 27, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
toc: true
language: 
  title-block-published: "Latest version"
---

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(lubridate)
library(tidybayes)
library(marginaleffects)
library(patchwork)
library(targets)

@@

tar_load(prelim_plot_data)
tar_load(h1_plot_data)
tar_load(h2_plot_data)

# Plotting functions
invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
```

Combating a deadly virus like SARS-CoV-2 requires extraordinary public health measures. In an effort to fight the Covid-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries are ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights principles. Some human rights treaties recognize states' needs for emergency measures when confronted with public health concerns and contain escape clauses that establish safeguards to prevent permanent erosion of these rights during emergencies. This temporary suspension of state compliance with its international treaty provisions occurs through a process known as derogations. However, the abuse of emergency measures during these derogations raises concerns about the long-lasting impact of these measures on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that in May 2023 it would be ending the emergency measures instated at the beginning of the pandemic—a move that that has occurred in many other countries as well. With the lifting of these measures, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this period. What did derogations mean for compliance? Are derogating and non-derogating states different in terms of how they implemented emergency measures during Covid-19? Were derogating states more likely to exploit the provisions and to abuse human rights during Covid-19? These questions are important for the study of international law, compliance, and times of crises.

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, the CoronaNet Covid-19 Government Response Event, and United Nations treaty derogation data we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. Our main takeaways are that derogating states did largely what they signaled they would through the derogations. States that derogated and specified time limits to the emergency provisions generally stuck to the time limits and did implement the emergency practices they signaled they would. Additionally, we find that, on average, derogating states are less likely to abuse emergency measures. On the other hand, states that did not derogate from their international human rights obligations and still impose emergency measures are more likely impose discriminatory policies, violate non-derogable rights such right to life AND WHAT ELSE, and impose emergency measures without time limits. Explain this result more - ANDREW "Civil society repression increases probability of abusive enforcement and discriminatory policies; derogations mitigate that probability."

@@

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. But we know much less about how international treaty provisions intended to safeguard rights might actually worsen human rights outcomes. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021].
 EXPAND ON RESULT In this scenario, when international factors may not be enough to constrain state actions, assessing the effect of domestic-level factors such as a robust civil society that potentially can mitigate the adverse effects of derogations becomes essential. WE FIND.

@@

Below, we first explore the purpose of derogations during the pandemic. In doing so, we focus specifically on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We then assess states' behavior towards civil society and these organizations' role in mitigating the adverse effects of emergency measures. We conclude with the implications' of our findings for the study of international law, times of crises, and the role of civil society.

@@

Some states new to derogations introducing them during the COVID-19 pandemic [@Comstock2023]. However, past treaty action behavior may help explain entering derogations during this time. To examine this first step, we code two variables 1) past ICCPR derogation behavior and 2) past ICCPR post-commitment action behavior. States receive a 1 if they engaged in this legal behavior before 2020 and 0 if they did not. This will allow us to empirically test for the first step of engagement before analyzing any impact COVID-19 era derogations had on different rights practices. There were 36 states that submitted ICCPR derogations and 60 that submitted any type of post-commitment action between 1966-2019.

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to more broadly examine the relationship between treaty derogations, civil society, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl:data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford Covid-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay at home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

@@

We examine the relationship between formal treaty derogations and different human rights and policy outcomes in three stages (see @tbl:vars). First, to explore whether derogations were associated with emergency state responses in general, we look at the connection between ICCPR derogations and five different emergency public health measures recorded by the Oxford COVID-19 Government Response Tracker [@HaleAngristGoldszmidt:2021]: (1) cancelling public events, (2) imposing restrictions on gathering, (3) closing public transit, (4) restricting internal movement, and (5) limiting international travel. Each of these outcomes is recorded as an ordered categorical variable, with levels ranging from "no measures" to increasing counts and severity of emergency measures. As we are most interested in the overall association between derogations and emergency measures, we collapse these variables into binary outcomes that indicate whether a country had no emergency measures or some emergency measures in a given week.

Next, to test our first hypothesis, we explore the relationship between formal treaty derogations on four different human rights and four policy outcomes from V-Dem's PanDem project [@vdem-pandem]: (1) whether pandemic-related emergency measures violated the ICCPR's principle of non-discrimination, (2) whether countries derogated from non-derogable ICCPR rights, (3) whether emergency measures failed to state a time limit, and (4) whether state security forces violated physical integrity rights while enforcing emergency measures. Similar to the Oxford data, each of these outcomes is an ordered category measuring the severity of the outcomes, including no, minor, moderate, and major violations. We do not collapse these variables into binary indicators and instead work with them as ordered categories. Finally, we test our second hypothesis by exploring the more complex relationship treaty derogations, civil society regulations, and four policy outcomes from the PanDem project. We look at the same four policies as the first hypothesis with one exception: since civil society is often concerned with advocacy and information provision during public emergencies, we replace whether emergency measures have a time limit with an ordinal measure of whether limits were placed on media freedom as part of a state's pandemic response. Importantly, since these PanDem variables are all measured quarterly, while our main data is recorded weekly, these variables act more like country-level fixed effects in our models, capturing the overall level of the severity of pandemic response each week.

@@

Because these models provide coefficients on the logged odds scale, and because the ordered logistic models require the incorporation of coefficients for each threshold to correctly interpret, we do not discuss the results of the models using raw model estimates (though we include complete regression tables in the appendix). Instead we rely on probability-scale predictions and marginal effects whenever possible. We use median values from our models' posterior distributions as point estimates and provide credible intervals using the 95% highest posterior density. For hypothesis testing, we report the posterior probability that predictions and marginal effects are above or below zero.

@@

December 2020 special issue - called for people to do more descriptive stuff on covid restrictions

@@

Based on our model results, derogations have different ASDF ASDF

There is no substantial difference between derogating and non-derogating countries when considering the probability of imposing discriminatory policies or engaging in abusive enforcement. Both types of countries follow nearly identical trends over the first 15 months of the pandemic, with an 85% predicted probability of no discriminatory policy in March 2020 rising to nearly 100% by June 2021. The next most probable outcomes—minor and moderate levels of discriminatory policies—are equally likely regardless of derogation status.

In contrast, the patterns in predicted probabilities of the adherence to time limit measures are influenced strongly by derogation status. For non-derogating countries, the two most likely predicted outcomes throughout the beginning of the pandemic are to have no issues with time limits (with a roughly 60% probability) and to have moderate issues with limits, or an absence of an end date for emergency measures (with 30–40% probability). The middle ground of only minor issues with time limits is exceptionally unlikely. Countries that formally derogated, on the other hand, have a 90% probability of having no issues with time limits throughout the 15 months, and only a less-than-10% probability of moderate issues. The derogation process requires that emergency measures be time-limited, and countries that care about their treaty obligations appear to actually follow the time limits they establish.

TODO: Kind of help with non derogable

Derogations therefore appear to help protect against limitless emergency measures and somewhat against violations of non-derogable rights, but they don't do much to protect against actual observed human rights abuses like discriminatory policies and abuse enforcement of those policies. They help with bureaucratic processes, not the implementation of those processes.

@@

h1_plots <- h1_plot_data %>% 
  # Predicted probabilties
  mutate(pred_plot = pmap(list(pred_data, nice, family), \(.data, .nice_name, .family) {
    if (.family == "cumulative") {
      .data %>% 
        ggplot(aes(
          x = year_week_day, y = .epred, ymin = .lower, ymax = .upper,
          color = .category, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        scale_color_manual(values = c(clrs[c(2, 4, 6, 7)])) +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        scale_y_continuous(labels = label_percent()) +
        labs(x = NULL, y = "Predicted probability", 
             color = "Severity", fill = "Severity",
             title = .nice_name) +
        facet_wrap(vars(derogation_ineffect)) +
        theme_pandem()
    } else {
      .data %>% 
        ggplot(aes(
          x = year_week_day, y = .epred, ymin = .lower, ymax = .upper,
          color = derogation_ineffect, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        scale_color_manual(values = c(clrs[1], clrs[8])) +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        scale_y_continuous(labels = label_percent()) +
        labs(x = NULL, y = "Predicted probability", 
             color = "Derogation in effect", fill = "Derogation in effect",
             title = .nice_name) +
        theme_pandem()
    }
  })) %>% 
  # Marginal effects / contrasts
  mutate(mfx_plot = pmap(list(mfx_data, nice, family), \(.data, .nice_name, .family) {
    if (.family == "cumulative") {
      .data %>% 
        mutate(across(c(draw, .lower, .upper), ~ . * 100)) %>%
        ggplot(aes(
          x = year_week_day, y = draw, ymin = .lower, ymax = .upper,
          color = group, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        geom_hline(yintercept = 0, linewidth = 0.5, color = "grey50") +
        scale_color_manual(values = c(clrs[c(2, 4, 6, 7)]), guide = "none") +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        labs(x = NULL,
             y = "Percentage point difference\nin probability of outcome level",
             title = .nice_name) +
        facet_wrap(vars(group), ncol = 2) +
        theme_pandem()
    } else {
      .data %>% 
        mutate(across(c(draw, .lower, .upper), ~ . * 100)) %>%
        ggplot(aes(
          x = year_week_day, y = draw, ymin = .lower, ymax = .upper,
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3, fill = clrs[5], color = clrs[5]) +
        geom_hline(yintercept = 0, linewidth = 0.5, color = "grey50") +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        labs(x = NULL,
             y = "Percentage point difference\nin probability of outcome",
             title = .nice_name) +
        theme_pandem()
    }
  }))

@@

Civil society repression, on the other hand, adds an additional wrinkle to the story

The derogations vs. non-derogations split matters when looking across the range of civil society repression. In countries with bad repression that don't derogate, there's a higher likelihood of discriminatory policies and abusive enforcement. Countries with similar levels of civil society repression that *do* formally derogate, however, have a nearly 100% predicted probability having no discriminatory policies. The derogation process could thus serve as a strong signal that emergency measures are indeed planned to be limited and non-discriminatory even in countries where civil society is under threat.

@@

TODO: Abusive enforcement

TODO: Not much with limiting media

TODO: Nothing at all with nonderogable rights

CS deals with more on-the-ground stuff and monitoring abuses and shaming them - CS deals with more of the implementation of

On paper vs. in practice

@@

This research note examines states' responses to their international human rights treaty obligations, particularly the ICCPR, during the Covid-19 pandemic and its subsequent impact on human rights violations.

@@

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. However, the abuse of emergency derogations by many governments raises concerns about the long-lasting impact of these sweeping measures on human rights. Does a robust civil society mitigate the adverse effects arising from international treaty derogations? Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run. Using data from the Varieties of Democracy PanDem dataset and the Oxford Covid-19 Government Response Tracker, we first assess the variation in human rights responses among governments during the pandemic. We then assess the role of civil society in keeping governments accountable and mitigating the effects of policies that violate human rights. Our results hold important implications for understanding government compliance with international treaties and shedding light on how states of emergency may directly and indirectly lead to democratic backsliding.
date: March 5, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
link-citations: true
format: 
  html:
    appendix-style: plain
    toc: true
  pdf:
    mainfont: "IBM Plex Serif"
    sansfont: "Noto Sans"
    linestretch: 1.25
    toc: false
language: 
  title-block-published: "Latest version"
---

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(lubridate)
library(tidybayes)
library(marginaleffects)
library(patchwork)
library(colorspace)
library(modelsummary)
library(kableExtra)
library(targets)

@@

tar_load(c(prelim_plot_data, h1_plot_data, h2_plot_data))
tar_load(c(models_tbl_prelim, models_tbl_h1, models_tbl_h2))

# Plotting and modelsummary functions
invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
invisible(list2env(tar_read(modelsummary_functions), .GlobalEnv))
```

Combating a deadly virus like SARS-CoV-2 requires extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries are ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights principles. Some human rights treaties recognize states' needs for emergency measures when confronted with public health concerns and contain escape clauses that establish safeguards to prevent permanent erosion of these rights during emergencies. This temporary suspension of state compliance with its international treaty provisions occurs through a process known as derogations. However, the abuse of emergency measures during these derogations raises concerns about the long-lasting impact of these measures on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that in May 2023 it would be ending the emergency measures instated at the beginning of the pandemic—a move that that has occurred in many other countries as well. With the lifting of these measures, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this period. What did derogations mean for compliance? Are derogating and non-derogating states different in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit the provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and times of crises.

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. Our main takeaways are that derogating states did largely what they signaled they would through the derogations. States that derogated and specified time limits to the emergency provisions generally stuck to the time limits and did implement the emergency practices they signaled they would. Additionally, we find that, on average, derogating states are less likely to abuse emergency measures. On the other hand, states that did not derogate from their international human rights obligations and still impose emergency measures are more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits. We also find that civil society can play an important mitigating role in curbing discriminatory policies. States with low levels of civil society repression are the least likely to abuse human rights when enforcing emergency policies. The likelihood of discriminatory policies increase as civil society repression worsens, but derogations appear to protect against worsening abuses—even in countries with severe civil society repression, those that derogate have a nearly 100% probability of no discriminatory policies.

@@

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. But we know much less about how international treaty provisions intended to safeguard rights might actually worsen human rights outcomes. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. TODO: EXPAND ON RESULT In this scenario, when international factors may not be enough to constrain state actions, assessing the effect of domestic-level factors such as a robust civil society that potentially can mitigate the adverse effects of derogations becomes essential. TODO: WE FIND.

@@

Below, we first explore the purpose of derogations during the pandemic. In doing so, we focus specifically on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We then assess states' behavior towards civil society and these organizations' role in mitigating the adverse effects of emergency measures. We conclude with the implications of our findings for the study of international law, times of crises, and the role of civil society.

@@

Some states new to derogations introducing them during the COVID-19 pandemic [@Comstock:2023]. However, past treaty action behavior may help explain entering derogations during this time. To examine this first step, we code two variables 1) past ICCPR derogation behavior and 2) past ICCPR post-commitment action behavior. States receive a 1 if they engaged in this legal behavior before 2020 and 0 if they did not. This will allow us to empirically test for the first step of engagement before analyzing any impact COVID-19 era derogations had on different rights practices. There were 36 states that submitted ICCPR derogations and 60 that submitted any type of post-commitment action between 1966-2019.

@@

While some studies on repression, human rights, and the pandemic have focused within single country cases or across a region, we examine a global database to more broadly examine the relationship between treaty derogations, civil society, and state actions during the pandemic. We explore our hypotheses using data on COVID-19 derogations, state actions, health trends, and civil society measures that we collected from several different sources (see @tbl-data-sources), including start and end dates of ICCPR derogations [@UnitedNations:2023], formal government responses from the Oxford Covid-19 Government Response Tracker [@HaleAngristGoldszmidt:2021], indicators of the severity of government responses from the Varieties of Democracy project [@vdem-v12; @vdem-pandem], and COVID case and death counts from the World Health Organization's COVID-19 dashboard [@WHO:2023]. We include data from `r info$n_countries` countries during the first 15 months (or `r info$n_weeks` weeks) of the pandemic, from March 11, 2020 to June 30, 2021. While the Oxford Government Response Tracker provides daily data on a variety of government actions, submitting formal treaty responses to the ICCPR is less likely to be as immediate of an action as implementing domestic emergency actions like stay at home orders. Accordingly, we collapse this daily data into weekly totals, resulting in a panel with `r info$n_rows` country-week observations.

```{r tbl-data-sources}
#| tbl-cap: Sources for outcomes and explanatory variables
#| include: true
#| echo: false
tribble(
  ~`Outcome variables`,                        ~Description,                                                        ~Frequency,  ~Source,
  "Emergency public health measures",          "Binary indicator",                                                  "Weekly",    "Oxford Covid-19 Government Response Tracker",
  "Human rights and policy outcomes",          "Ordered categories for the severity of abuses",                     "Quarterly", "Varieties of Democracy PanDem",
  "ICCPR derogation",                          "Binary indicator",                                                  "Weekly",    "UN Treaty Collection",
  "Civil society repression",                  "≈ −3 to 3; higher values indicate less repression",                 "Annual",    "Varieties of Democracy",
  "New and cumulative COVID deaths and cases", "Counts",                                                            "Weekly",    "World Health Organization",
  "Past ICCPR derogation or action",           "Binary indicator",                                                  "Weekly",    "UN Treaty Collection",
  "Rule of law index",                         "0–1; higher values represent greater respect for rule of law",      "Annual",    "Varieties of Democracy",
  "Civil liberties index",                     "0–1; higher values represent greater respect for civil liberties",  "Annual",    "Varieties of Democracy",
  "Core civil society index",                  "0–1; higher values represent better environment for civil society", "Annual",    "Varieties of Democracy"
) %>% 
  kbl(booktabs = TRUE) %>% 
  kable_styling() %>% 
  pack_rows(index = c(" " = 2, "Primary explanatory variables" = 2, "Secondary explanatory variables" = 5),
            indent = FALSE)

```

@@

We examine the relationship between formal treaty derogations and different human rights and policy outcomes in three stages (see @tbl-vars). First, to explore whether derogations were associated with emergency state responses in general, we look at the connection between ICCPR derogations and five different emergency public health measures recorded by the Oxford Covid-19 Government Response Tracker [@HaleAngristGoldszmidt:2021]: (1) cancelling public events, (2) imposing restrictions on gathering, (3) closing public transit, (4) restricting internal movement, and (5) limiting international travel. Each of these outcomes is recorded as an ordered categorical variable, with levels ranging from "no measures" to increasing counts and severity of emergency measures. As we are most interested in the overall association between derogations and emergency measures, we collapse these variables into binary outcomes that indicate whether a country had no emergency measures or some emergency measures in a given week.

| General question                                    | Outcomes                                                                                                                                  | Main explanatory variable                   | Other explanatory variables                                                                                                                                              |
|----------------|------------------|----------------|----------------------|
| Derogations and emergency measures                  | Cancelling public events, Gathering restrictions, Close public transit, Restrictions on internal movement, Limits on international travel | ICCPR derogation in effect                  | New and cumulative COVID cases, new and cumulative COVID deaths, prior ICCPR derogations and actions                                                                     |
| Derogations and human rights (H~1~)                 | Discriminatory policy, Violations of non-derogable rights, No time limit measures, Abusive enforcement                                    | ICCPR derogation in effect                  | New and cumulative COVID cases, new and cumulative COVID deaths, prior ICCPR derogations and actions, rule of law index, civil liberties index, core civil society index |
| Derogations, human rights, and civil society (H~2~) | Discriminatory policy, Violations of non-derogable rights, Limiting media, Abusive enforcement                                            | ICCPR derogation × civil society repression | New and cumulative COVID cases, new and cumulative COVID deaths, prior ICCPR derogations and actions, rule of law index, civil liberties index                           |

: Overview of outcomes and explanatory variables included in models {#tbl-vars}

Next, to test our first hypothesis, we explore the relationship between formal treaty derogations on four different human rights and policy outcomes from V-Dem's PanDem project [@vdem-pandem]: (1) whether pandemic-related emergency measures violated the ICCPR's principle of non-discrimination, (2) whether countries derogated from non-derogable ICCPR rights, (3) whether emergency measures failed to state a time limit, and (4) whether state security forces violated physical integrity rights while enforcing emergency measures. Similar to the Oxford data, each of these outcomes is an ordered category measuring the severity of the outcomes, including no, minor, moderate, and major violations. We do not collapse these variables into binary indicators and instead work with them as ordered categories. Finally, we test our second hypothesis by exploring the more complex relationship treaty derogations, civil society regulations, and four policy outcomes from the PanDem project. We look at the same four policies as the first hypothesis with one exception: since civil society is often concerned with advocacy and information provision during public emergencies, we replace whether emergency measures have a time limit with an ordinal measure of whether limits were placed on media freedom as part of a state's pandemic response. Importantly, since these PanDem variables are all measured quarterly, while our main data is recorded weekly, these variables act more like country-level fixed effects in our models, capturing the overall level of the severity of pandemic response each week.

@@

Because these models provide coefficients on the logged odds scale, and because the ordered logistic models require the incorporation of coefficients for each threshold to correctly interpret, we do not discuss the results of the models using raw model estimates (though we include complete regression tables in the appendix). Instead we rely on probability-scale predictions and marginal effects whenever possible. To do so, we calculate predicted probabilities holding all explanatory variables at their typical values (median or mode). For logistic models, we estimate the probability of having a specific policy or violating a specific human right; for ordered models, we estimate the probability of each possible level of severity. To determine the effect of derogations, we calculate the contrast in predicted probabilities when derogation is set to true vs. when it is set to false (e.g., in ordered logistic models we compare the probability of major discrimination under derogation with the probability of major discrimination without derogation). We use median values from our models' posterior distributions as point estimates and provide credible intervals using the 95% highest posterior density. For hypothesis testing, we report the posterior probability that the between-level predicted differences are above or below zero.

@@

@@

The results of our models show that derogations have some protective effect against more bureaucratic forms of human rights abuses and less against practical abuses (see @fig-preds-mfx-h1). There is no substantial difference between derogating and non-derogating countries when considering the probability of imposing discriminatory policies or engaging in abusive enforcement. Both types of countries follow nearly identical trends over the first 15 months of the pandemic, with an 85% predicted probability of no discriminatory policy in March 2020 rising to nearly 100% by June 2021. The next most probable outcomes—minor and moderate levels of discriminatory policies—are equally likely regardless of derogation status. Moreover, countries that derogate were substantially less likely to engage in no abusive enforcement and instead turn to minor or moderate levels of abusive enforcement, particularly in the early months of the pandemic. Minor and moderate levels of excessive enforcement are each several percentage points more likely for derogating countries than for their non-derogating counterparts. The mechanisms behind this seeming reversal—where countries that follow correct human rights procedures end up enforcing emergency policies in a more abusive way—are unclear.

In contrast, the patterns in predicted probabilities of the adherence to time limit measures are influenced strongly by derogation status. For non-derogating countries, the two most likely predicted outcomes throughout the beginning of the pandemic are to have no issues with time limits (with a roughly 60% probability) and to have moderate issues with limits, or an absence of an end date for emergency measures (with 30–40% probability). The middle ground of only minor issues with time limits is exceptionally unlikely. Countries that formally derogated, on the other hand, have a 90% probability of having no issues with time limits throughout the 15 months, and only a less-than-10% probability of moderate issues. The derogation process requires that emergency measures be time-limited, and countries that care about their treaty obligations appear to actually follow the time limits they establish. Additionally, the risk of violating non-derogable rights is generally slightly lower in non-derogating states—countries that derogate are a percentage point less likely to violate non-derogable rights ($p(\Delta < 0) \approx 0.83$)

Derogations therefore appear to help protect against limitless emergency measures and somewhat against violations of non-derogable rights, but they do not appear to do much to protect against actual observed human rights abuses like discriminatory policies and abuse enforcement of those policies. We thus find strong support only for H~1c~—that derogating states are less likely to impose measures with no time limits. We find weak support for our hypothesis that derogating states will be less likely to violated non-derogable rights (H~1b~) and no support for our hypothesis that derogations protect against discriminatory policies (H~1c~). Derogations seem help with bureaucratic processes and might protect against severe violations of human rights, but they do not appear to influence the overall implementation of emergency measures.

@@

h1_plots <- h1_plot_data %>% 
  # Predicted probabilities
  mutate(pred_plot = pmap(list(pred_data, nice, family), \(.data, .nice_name, .family) {
    if (.family == "cumulative") {
      .data %>% 
        ggplot(aes(
          x = year_week_day, y = .epred, ymin = .lower, ymax = .upper,
          color = .category, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        scale_color_manual(values = c(clrs[c(2, 4, 6, 7)])) +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        scale_y_continuous(labels = label_percent()) +
        labs(x = NULL, y = "Predicted probability", 
             color = "Severity", fill = "Severity",
             title = .nice_name) +
        facet_wrap(vars(derogation_ineffect)) +
        theme_pandem()
    } else {
      .data %>% 
        ggplot(aes(
          x = year_week_day, y = .epred, ymin = .lower, ymax = .upper,
          color = derogation_ineffect, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        scale_color_manual(values = c(clrs[1], clrs[8])) +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        scale_y_continuous(labels = label_percent()) +
        labs(x = NULL, y = "Predicted probability", 
             color = "Derogation in effect", fill = "Derogation in effect",
             title = .nice_name) +
        theme_pandem()
    }
  })) %>% 
  # Marginal effects / contrasts
  mutate(mfx_plot = pmap(list(mfx_data, nice, family), \(.data, .nice_name, .family) {
    if (.family == "cumulative") {
      .data %>% 
        mutate(across(c(draw, .lower, .upper), ~ . * 100)) %>%
        ggplot(aes(
          x = year_week_day, y = draw, ymin = .lower, ymax = .upper,
          color = group, fill = after_scale(color),
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3) +
        geom_hline(yintercept = 0, linewidth = 0.5, color = "grey50") +
        scale_color_manual(values = darken(clrs[c(2, 4, 6, 7)], 0.5, space = "combined"), guide = "none") +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        labs(x = NULL,
             y = "Percentage point difference\nin probability of outcome level",
             title = .nice_name) +
        facet_wrap(vars(group), ncol = 2) +
        theme_pandem()
    } else {
      .data %>% 
        mutate(across(c(draw, .lower, .upper), ~ . * 100)) %>%
        ggplot(aes(
          x = year_week_day, y = draw, ymin = .lower, ymax = .upper,
          group = forcats::fct_rev(ordered(.width))
        )) +
        geom_lineribbon(alpha = 0.3, fill = clrs[5], color = clrs[5]) +
        geom_hline(yintercept = 0, linewidth = 0.5, color = "grey50") +
        scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
        labs(x = NULL,
             y = "Percentage point difference\nin probability of outcome",
             title = .nice_name) +
        theme_pandem()
    }
  }))

@@

Civil society repression, on the other hand, adds an additional wrinkle to the story. Countries with severe level of repression (i.e. negative repression scores) that do not derogate have a 25% predicted probability of minor to major discriminatory policies, while similarly repressive countries that do derogate have nearly a 0% predicted probability of discriminatory policies (see @fig-preds-mfx-h2). This represents a substantial 5–10 percentage point difference between derogating and non-derogating countries with repressive civil societies. The derogation process could thus serve as a strong signal that emergency measures are indeed planned to be limited and non-discriminatory even in countries where civil society is under threat.

However, there are no other substantial differences in predicted probabilities across derogation status for another other outcomes. Derogating and non-derogating states are generally equally likely to violate non-derogable rights (H~2a~), limit media (H~2d~), and engage in abusive enforcement (H~2c~) regardless of their level of civil society repression. We thus only find support for H~2b~—if states are repressive toward civil society organizations, they are more likely to use discriminatory policies, while derogations strongly negate this trend. This may be advocacy-focused civil society organizations in particular tend to deal with on-the-ground monitoring, advocacy, and shaming, and is more likely to call out governments for unequal and unjust implementation of their emergency policies.

@@

@@

This research note examines states' responses to their international human rights treaty obligations, particularly the ICCPR, during the COVID-19 pandemic and its subsequent impact on human rights violations.

@@

```{r tbl-results-full-prelim}
#| tbl-cap: Complete results from models showing relationship between derogations and emergency policies
#| include: true
#| echo: false
notes <- c(
  "Estimates are median posterior log odds from ordered logistic and binary logistic regression models; 95% credible intervals (highest density posterior interval, or HDPI) in brackets.", 
  "Total \\(R^2\\) considers the variance of both population and group effects;", 
  "marginal \\(R^2\\) only takes population effects into account."
)

modelsummary(models_tbl_prelim,
             estimate = "{estimate}",
             statistic = "[{conf.low}, {conf.high}]",
             coef_map = coef_map,
             gof_map = gof_map,
             output = "kableExtra",
             fmt =  list(estimate = 2, conf.low = 2, conf.high = 2)) %>% 
  kable_styling(htmltable_class = "table-sm light-border") %>% 
  footnote(general = notes, footnote_as_chunk = TRUE)
```
